---
title: "MovieLens Capstone Report"
author: "Yvonne Kirschler"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(caret)
library(lubridate)
```

# Introduction

This capstone project is part of the HarvardX Data Science Professional Certificate.  
The goal is to build a model that predicts movie ratings based on the MovieLens 10M dataset.  
Model performance is evaluated using RMSE.  
The final model is tested only once on `final_holdout_test` in accordance with project rules.

# Data Preparation

```{r data-prep}
if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")

library(tidyverse)
library(caret)

dl <- "ml-10M100K.zip"
if(!file.exists(dl)) {
  download.file("https://files.grouplens.org/datasets/movielens/ml-10m.zip", dl)
}
if(!file.exists("ml-10M100K/ratings.dat")) {
  unzip(dl)
}

ratings <- as.data.frame(str_split(read_lines("ml-10M100K/ratings.dat"), fixed("::"), simplify = TRUE),
                         stringsAsFactors = FALSE)
colnames(ratings) <- c("userId", "movieId", "rating", "timestamp")
ratings <- ratings %>%
  mutate(userId = as.integer(userId),
         movieId = as.integer(movieId),
         rating = as.numeric(rating),
         timestamp = as.integer(timestamp))

movies <- as.data.frame(str_split(read_lines("ml-10M100K/movies.dat"), fixed("::"), simplify = TRUE),
                        stringsAsFactors = FALSE)
colnames(movies) <- c("movieId", "title", "genres")
movies <- movies %>%
  mutate(movieId = as.integer(movieId))

movielens <- left_join(ratings, movies, by = "movieId")

set.seed(1)
test_index <- createDataPartition(y = movielens$rating, times = 1, p = 0.1, list = FALSE)
edx <- movielens[-test_index,]
temp <- movielens[test_index,]

final_holdout_test <- temp %>%
  semi_join(edx, by = "movieId") %>%
  semi_join(edx, by = "userId")

removed <- anti_join(temp, final_holdout_test)
edx <- rbind(edx, removed)

rm(dl, ratings, movies, test_index, temp, movielens, removed)
```

# Exploratory Data Analysis

```{r eda}
glimpse(edx)
dim(edx)

n_users <- n_distinct(edx$userId)
n_movies <- n_distinct(edx$movieId)

print(n_users)
print(n_movies)

plot_rating_dist <- edx %>%
  count(rating) %>%
  arrange(desc(n)) %>%
  ggplot(aes(x = rating, y = n)) +
  geom_col(fill = "steelblue") +
  labs(title = "Rating distribution", x = "Rating", y = "Count") +
  theme_minimal()

print(plot_rating_dist)

top_movies <- edx %>%
  group_by(title) %>%
  summarize(count = n()) %>%
  arrange(desc(count)) %>%
  slice(1:10)

print(top_movies)

top_genres <- edx %>%
  separate_rows(genres, sep = "\\|") %>%
  count(genres, sort = TRUE) %>%
  slice(1:10)

print(top_genres)

plot_time <- edx %>%
  mutate(date = as_datetime(timestamp)) %>%
  ggplot(aes(x = date)) +
  geom_histogram(binwidth = 30*24*60*60, fill = "darkgreen", color = "white") +
  labs(title = "Ratings over time", x = "Date", y = "Count")

print(plot_time)

plot_user_activity <- edx %>%
  count(userId) %>%
  ggplot(aes(x = n)) +
  geom_histogram(bins = 30, fill = "purple") +
  labs(title = "Number of ratings per user", x = "Ratings", y = "Users")

print(plot_user_activity)
```

# Movie Effect Model

```{r model-movie}
mu <- mean(edx$rating)

movie_avgs <- edx %>%
  group_by(movieId) %>%
  summarize(b_i = mean(rating - mu))

predicted_ratings <- final_holdout_test %>%
  left_join(movie_avgs, by = "movieId") %>%
  mutate(pred = mu + b_i)

rmse_movie_effect <- sqrt(mean((predicted_ratings$rating - predicted_ratings$pred)^2))

print(rmse_movie_effect)
```

# Movie + User Effect Model

```{r model-user}
movie_avgs <- edx %>%
  group_by(movieId) %>%
  summarize(b_i = mean(rating - mu))

user_avgs <- edx %>%
  left_join(movie_avgs, by = "movieId") %>%
  group_by(userId) %>%
  summarize(b_u = mean(rating - mu - b_i))

predicted_ratings <- final_holdout_test %>%
  left_join(movie_avgs, by = "movieId") %>%
  left_join(user_avgs, by = "userId") %>%
  mutate(pred = mu + b_i + b_u)

rmse_user_effect <- sqrt(mean((predicted_ratings$rating - predicted_ratings$pred)^2))

print(rmse_user_effect)
```

# Regularized Model

```{r model-regularized}
lambdas <- seq(0, 10, 0.25)

RMSE <- function(true_ratings, predicted_ratings) {
  sqrt(mean((true_ratings - predicted_ratings)^2))
}

rmse_results <- sapply(lambdas, function(lambda) {
  b_i <- edx %>%
    group_by(movieId) %>%
    summarize(b_i = sum(rating - mu) / (n() + lambda), .groups = "drop")
  
  b_u <- edx %>%
    left_join(b_i, by = "movieId") %>%
    group_by(userId) %>%
    summarize(b_u = sum(rating - mu - b_i) / (n() + lambda), .groups = "drop")
  
  predicted_ratings <- final_holdout_test %>%
    left_join(b_i, by = "movieId") %>%
    left_join(b_u, by = "userId") %>%
    mutate(
      b_i = coalesce(b_i, 0),
      b_u = coalesce(b_u, 0),
      pred = mu + b_i + b_u
    ) %>%
    pull(pred)
  
  RMSE(final_holdout_test$rating, predicted_ratings)
}, simplify = TRUE)

rmse_results_df <- data.frame(lambda = lambdas, rmse = rmse_results)

print(rmse_results_df)

best_result <- rmse_results_df %>% arrange(rmse) %>% slice(1)
print(best_result)
```

# Final Model & RMSE

```{r final-model}
lambda <- 5.25

b_i <- edx %>%
  group_by(movieId) %>%
  summarize(b_i = sum(rating - mu) / (n() + lambda), .groups = "drop")

b_u <- edx %>%
  left_join(b_i, by = "movieId") %>%
  group_by(userId) %>%
  summarize(b_u = sum(rating - mu - b_i) / (n() + lambda), .groups = "drop")

final_predictions <- final_holdout_test %>%
  left_join(b_i, by = "movieId") %>%
  left_join(b_u, by = "userId") %>%
  mutate(
    b_i = coalesce(b_i, 0),
    b_u = coalesce(b_u, 0),
    pred = mu + b_i + b_u
  )

final_rmse <- sqrt(mean((final_predictions$rating - final_predictions$pred)^2))

print(final_rmse)
```

# RMSE Summary

```{r rmse-summary}
data.frame(
  Model = c("Movie Effect", "Movie + User Effect", "Regularized Final"),
  RMSE = c(rmse_movie_effect, rmse_user_effect, final_rmse)
)
```

# Conclusion

The final model achieved an RMSE of `r round(final_rmse, 6)`, using regularized Movie + User effects and lambda tuning (`lambda = 5.25`).  
This model performs better than the simpler baselines and meets the capstone criteria.